---
title: "Chamala_week12_Final project"
editor_options:
  chunk_output_type: inline
date: "02/22/20"
output:
  pdf_document: default
  word_document: default
---

```{r}
#installing required packages
library(dplyr)
library(utils)
library(ggplot2)
library(tidyverse)
library(readxl)
library(openxlsx)
library(data.table)
library(tidyr)
library(zoo)
library(plyr)
library(janitor)
library(base)

#	Children Under 3 yrs with Elevated Blood Lead Levels (BLL)
Age3 <- read.csv("Children_Tested_for_Lead_by_Age_3.csv", header = TRUE)


#	Children Under 6 yrs with Elevated Blood Lead Levels (BLL)
Age6 <- read.csv("Children_Under_6_yrs_with_Elevated_Blood_Lead_Levels__BLL_.csv", header = TRUE)


#	CDC National Childhood Blood Lead Surveillance- Children under 6 yrs with Elevated Blood Lead Levels (BLL)
CBLS_National_Table_Update_042619 <- read_excel("CBLS-National-Table-Update-042619.xlsx", sheet = "By State and Year", col_names = FALSE)

```


Cleaning CBLS_National_Table_Update_042619 file

```{r}
#Deleting unwanted rows
national <- CBLS_National_Table_Update_042619[-c(1:4,323:326),  ]

#Assigning new headers
names(national) <- c("year","state", "Total_population<72m", "Children_tested<72m", "%children_tested<72m", "confirmed_BLLs>5µg/dL", "%confirmed_BLLs>5µg/dL", "confirmed_BLLs>10µg/dL", "%confirmed_BLLs>10µg/dL", "confirmed_BLLs_grp_5-9 µg/dL", "confirmed_BLLs_grp_10-14 µg/dL", "confirmed_BLLs_grp_15-19 µg/dL", "confirmed_BLLs_grp_20-24 µg/dL", "confirmed_BLLs_grp_25-44 µg/dL", "confirmed_BLLs_grp_>45 µg/dL" )

national$state <- na.locf(national$state)

#filling NA values with 0
national[national=="N/A"] <- 0

#Deleting data which are entirely empty
national = filter(national, !(national$state %in% c("Alaska", "Arkansas", "California", "Florida", "Hawaii", "Idaho", "Iowa", "Kansas", "Montana", "Nebraska", "Nevada", "New York City", "North Dakota", "Oregon", "South Carolina", "South Dakota", "Texas", "Utah", "Virginia", "Wyoming")))

national[, c(1,3:15)] <- sapply(national[, c(1,3:15)], as.numeric)

```

Checking for missing data 

```{r}
any(is.na(Age3))
any(is.na(Age6))
any(is.na(national))

```
#Removing columns that are empty 
For the Age6 data set, I had to use a different strategy as I do not need the Notes columns. I could not use the "remove_empty" function because, they are not all empty. There are some generic notes in some of the columns so I had to take a different approach to remove columns that all end with the string - NOTES


```{r}
Age3_new <- remove_empty(Age3, which = c("cols"))
head(Age3_new)

# Removing all the columns that end with NOTES 
Age6_new <- Age6 %>% select(-contains("NOTES"))
head(Age6_new)
```



# Renaming long column names with shorter ones for Age 3 and Age 6 data sets
```{r setup, include=TRUE, echo=FALSE}

names(Age3_new) <- c("geo_type", "geo_area_id", "geo_area_name", "borough_id", "time_period", "age3_number", "age3_percentage")

names(Age6_new) <- c("geo_type", "geo_area_id", "geo_area_name", "borough_id", "time_period", "age6_bll_5", "age6_bll_10", "age6_bll_15", "age6_bll_total", "age6_bll_5_per1000", "age6_bll_10_per1000", "age6_bll_15_per1000" )


```

Displaying cleaned data 

```{r setup, include=TRUE, echo=FALSE}
head(Age3_new)
head(Age6_new)
head(national)
```

Slicing data to create subsets for analysis
```{r setup, include=TRUE, echo=FALSE}
Age3_borough <- subset(Age3_new, Age3_new$geo_type == "Borough")

Age3_neighborhood <- subset(Age3_new, Age3_new$geo_type == "Neighborhood (UHF 42)")

Age6_borough <- subset(Age6_new, Age6_new$geo_type == "Borough")

Age6_neighborhood <- subset(Age6_new, Age6_new$geo_type == "Neighborhood (UHF 42)")

national_states <- subset(national[1:192, ])

national_USwide <- subset(national[193:198, ])

```

Visualizing distributions of various subsets created

```{r setup, include=TRUE, echo=FALSE}
#Bar plot for Age3 borough data set
Age3_borough$time_period <- as.factor(Age3_borough$time_period)
ggplot(Age3_borough, aes(fill=Age3_borough$geo_area_name, y=Age3_borough$age3_number, x=Age3_borough$time_period)) + 
    geom_col(position="dodge", stat="identity") + labs(x = "Time period", y = "Number of children tested positive", title = "Children tested positive under Age 3years")

#Bar plot for Age6 borough data set
Age6_borough$time_period <- as.factor(Age6_borough$time_period)
ggplot(Age6_borough, aes(fill=Age6_borough$geo_area_name, y=Age6_borough$age6_bll_total, x=Age6_borough$time_period)) + geom_bar(position="dodge", stat="identity") + labs(x = "Time period", y = "Number of children tested positive", title = "Children tested positive under Age 6years")

#scatter plot for Age3 neighborhood data set
ggplot(data = Age3_neighborhood, aes(x = Age3_neighborhood$geo_area_name, y = Age3_neighborhood$age3_number)) + geom_point(color = Age3_neighborhood$time_period, position = "jitter") + theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text=element_text(size=8), text=element_text(size=5)) + labs(x = "Neighborhood", y = "Number of children tested positive", title = "Children tested positive under Age 3years by borough")


#scatter plot for Age6 neighborhood data set
ggplot(data = Age6_neighborhood, aes(x = Age6_neighborhood$geo_area_name, y = Age6_neighborhood$age6_bll_total)) + geom_point(color = Age6_neighborhood$time_period, position = "jitter") + theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text=element_text(size=8), text=element_text(size=5)) + labs(x = "Neighborhood", y = "Number of children tested positive", title = "Children tested positive under Age 6years by borough")

#bar plot for National states data set
national_states$year <- as.factor((national_states$year))
ggplot(national_states, aes(fill=national_states$year, y=national_states$'Children_tested<72m', x=national_states$state, space =1)) + geom_bar(position="dodge", stat="identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text=element_text(size=8), text=element_text(size=5)) + labs(x = "State", y = "Number of children tested positive", title = "Children tested positive under Age 6years by state")

#bar plot for National US wide data set
national_USwide$year <- as.factor((national_USwide$year))
ggplot(national_USwide, aes(fill=national_USwide$year, y=national_USwide$'Children_tested<72m', x=national_USwide$state, space =1)) + geom_bar(position="dodge", stat="identity") + labs(x = "State", y = "Number of children tested positive", title = "Children tested positive under Age 6years by USwide")


```




```{r setup, include=TRUE, echo=FALSE}
library(corrplot)
#Selecting only numeric values to perform correlation analysis for Age6 dataset.
Age6_num_data <- Age6_new[, sapply(Age6_new, is.numeric)]
Age6_cor <- cor(Age6_num_data, use="pairwise.complete.obs", method = c("spearman"))

corrplot(Age6_cor)
         
ggplot(Age6_new, aes(x=Age6_new$time_period, y=Age6_new$age6_bll_5)) + geom_point(position = "jitter")        
```

